---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
# Introduction
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment can be downloaded from the course web site:

Dataset: Activity monitoring data [52K]
The variables included in this dataset are:

steps: Number of steps taking in a 5-minute interval (missing values are coded as 𝙽𝙰)
date: The date on which the measurement was taken in YYYY-MM-DD format
interval: Identifier for the 5-minute interval in which measurement was taken
The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

## R Markdown Preparation
Show the code as an output throughout.  Set the default of echo to true.

```{r R Markdown Prepration}
library(knitr)
opts_chunk$set(echo = TRUE)
opts_chunk$set(message = TRUE)
opts_chunk$set(fig.path = "PA1_template_figures/") 


```
## Load the R Packages
Preload the required R packages to meet the assignment objectives.
```{r Load the Packages, echo=TRUE}
library(stringr)
library(ggplot2)
library(lubridate)
library(dplyr)

```
## Loading and preprocessing the data
Show any code that is needed to  
1. Load the data (i.e. read.csv())  
2. Process/transform the data (if necessary) into a format suitable for your analysis

### Loading the Data
Download the data from the source URL and unzip.
1. Download from https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip
2. Unzip the zip file using unzip.
3. Load the activity data into a data frame using read.csv.


```{r Loading the Data, echo=TRUE}
# Download the zipped data from https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
targetFileName <- "repdata-data-activity.zip"
  
# Download only if the zip file hasn't been dowloaded yet.
if (!file.exists(targetFileName)) {
  download.file(url, targetFileName)  
}
  
## Unzip the files if they haven't been extracted
unzipFolderName <- "repdata-data-activity"
unzip(targetFileName, exdir = unzipFolderName)

activityData <- read.csv(paste0(unzipFolderName, "/activity.csv", sep = ""), stringsAsFactors = FALSE)
  
```
### Preprocessing the data
Format the data into the correct data types.
1. Convert the date column to a date type.
2. Add the datetime column.
3. Add the time column based on the interval column.
4. Omit NAs.

```{r Preprocessing the data, echo=TRUE}
activityData$intervalTimeFormat <- gsub("([0-9]{1,2})([0-9]{2})", "\\1:\\2", str_pad(activityData$interval, 4, side = "left", pad = "0"))

activityData$datetime <- strptime(paste0(activityData$date, activityData$intervalTimeFormat, sep=""), format="%Y-%m-%d %H:%M")
activityData$date <- ymd(activityData$date)
activityData$time <- hms(paste0(activityData$intervalTimeFormat, ":00", sep=""))
activityData$intervalTimeFormat <- NULL

```

Check the data with str() and head().
```{r Check the data with str() and head(), echo=TRUE}
str(activityData)
head(activityData)

```

## What is mean total number of steps taken per day?
For this part of the assignment, you can ignore the missing values in the dataset.

1. Calculate the total number of steps taken per day
2. If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day
3. Calculate and report the mean and median of the total number of steps taken per day

### Result
1. Calculate the total number of steps taken per day using dplyr and grouping by the date column. Note that using tapply and doing a sum shows 0 steps for 2012-10-01 and other dates. 

```{r Calculate Total Number of Steps Per Day, echo=TRUE}

totalSteps <- activityData[,c("steps", "date")] %>% 
              filter(!is.na(steps)) %>%
              group_by(date) %>%
              summarize(steps = sum(steps)) %>%
              print

```
2. Make a histogram of the total number of steps taken each day.  Use ggplot.

```{r , echo=TRUE}
###```{r histogram of the total number of steps taken each day, echo=TRUE}
g <- ggplot(totalSteps, aes(steps))
g <- g + geom_histogram(fill ="blue", binwidth = 1000) +
      guides(fill = F) +
      xlab("Steps per Day") +
      ylab("Frequency") +
      ggtitle('Histogram of Steps per Day') +
      theme(plot.title = element_text(hjust = 0.5), legend.position = "none")
print(g)

```

3. Calculate and report the mean and median of the total number of steps taken per day

```{r mean and median of the total number of steps taken per day, echo=TRUE}
meanSteps <- mean(totalSteps$steps, na.rm = TRUE)
medianSteps <- median(totalSteps$steps, na.rm = TRUE)
sprintf("Mean steps are %.0f and median steps are %.0f.", meanSteps, medianSteps)
```
## What is the average daily activity pattern?
1. Make a time series plot (i.e. type ="1𝚕") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

### Result
1. Using dplyr, calculate the average number of steps in each 5 minute interval per day

```{r Average Number of Steps for each 5 minute interval, echo=TRUE}
meanStepsPerInterval <- activityData[,c("steps", "interval")] %>%
                        filter(!is.na(steps)) %>%
                        group_by(interval) %>%
                        summarize(mean_steps = mean(steps)) %>%
                        print

```
Use ggplot for the making the time series plot.
```{r Time Series Plot Average Number of Steps for each 5 minute interval, echo=TRUE}
g <- ggplot(meanStepsPerInterval, aes(x=interval, y=mean_steps))
g <- g + geom_line(color ="blue") +
      guides(fill = F) +
      xlab("Interval") +
      ylab("Average Number of Steps") +
      ggtitle('Time Series Plot of the Average Number of Steps') +
      theme(plot.title = element_text(hjust = 0.5), legend.position = "none")
print(g)
```
2. Using which.max, find the 5 minute interval with the maximum number of average steps.

```{r 5 minute interval with the maximum number of average steps, echo=TRUE}
intervalMax <- meanStepsPerInterval[which.max(meanStepsPerInterval$steps),]
sprintf("5 Minutes Interval with the maximum number of average steps (%.0f) is %i.", intervalMax$steps, intervalMax$interval)

```

## Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs).
2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
4. Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

### Results
1.  Summarize the missing values.

```{r Total Number of Missing Values in the Activity Dataset, echo=TRUE}
sumStepsNA <- sum(is.na(activityData$steps))
sprintf("Total number of missing values in the activity dataset is %i", sumStepsNA)

```
2. For the impute strategy, if the number of steps is NA, use the mean for the interval.  If the mean for the day is still NA, use 0 steps.

```{r Substitute NA with the mean steps, echo=TRUE}
#Get the mean steps per day
meanStepsPerDay <- activityData[!is.na(activityData$steps),c("steps", "date")] %>% 
                    group_by(date) %>% 
                    summarize(mean_steps = mean(steps))

head(meanStepsPerDay, 10)
```
3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r Create new dataset filling in missing data, echo = TRUE}
#Merge the data sets activityData and meanStepsPerDay but do a left outer join.
#activityDataImpute <- merge(activityData, meanStepsPerDay, by = "date", all.x = TRUE, incomparables = 0)
activityDataImpute <- merge(activityData, meanStepsPerInterval, by = "interval", all.x = TRUE)

activityDataImpute$steps <- ifelse(!is.na(activityDataImpute$steps), activityDataImpute$steps, 
                                     ifelse(!is.na(activityDataImpute$mean_steps), activityDataImpute$mean_steps,
                                            0)
                                     )
# Remove the mean_step column
activityDataImpute <- activityDataImpute[, c("steps", "date", "interval", "time", "datetime")]   

head(activityDataImpute, 10)
```
4.  Impute - Histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day.

4a. Impute - Histogram of the total number of steps taken each day with missing steps filled in using mean steps per day.
    The histogram shows that the 0 step increased in frequency as there was data where the number of steps on given day is NA for all intervals and we assigned them with 0 steps.
```{r Impute histogram of the total number of steps taken each day, echo=TRUE}
totalStepsImpute <- activityDataImpute[,c("steps", "date")] %>% 
                    filter(!is.na(steps)) %>%
                    group_by(date) %>%
                    summarize(steps = sum(steps)) %>%
                    print

g <- ggplot(totalStepsImpute, aes(steps))
g <- g + geom_histogram(fill ="green", binwidth = 1000) +
      guides(fill = F) +
      xlab("Steps per Day") +
      ylab("Frequency") +
      ggtitle('Histogram of Steps per Day Imputed') +
      theme(plot.title = element_text(hjust = 0.5), legend.position = "none")
print(g)
           
```
4a. Non-Impute and Impute - Histogram of the total number of steps taken each day to quickly show the comparison.
    The result shows that using the mean affected the total number of steps taken per day.

```{r Non-Impute Impute histogram of the total number of steps taken each day, echo=TRUE}
gCombined <- ggplot() +
      geom_histogram(data=totalStepsImpute, aes(steps), fill ="green", binwidth = 1000, show.legend = TRUE) +
      geom_histogram(data=totalSteps, aes(steps), fill ="blue", binwidth = 1000, show.legend = TRUE) +
      guides(fill = T) +
      xlab("Steps per Day") +
      ylab("Frequency") +
      ggtitle('Histogram of Steps per Day - Non-Imputed and Imputed') +
      theme(plot.title = element_text(hjust = 0.5))
print(gCombined)
              
```

4b. Impute - Calculate and report the mean and median of the total number of steps taken per day

```{r Impute mean and median of the total number of steps taken per day, echo=TRUE}
meanStepsImpute <- mean(totalStepsImpute$steps, na.rm = TRUE)
medianStepsImpute <- median(totalStepsImpute$steps, na.rm = TRUE)
sprintf("For the dataset with filled in missing data, mean steps are %.0f and median steps are %.0f.", meanStepsImpute, medianStepsImpute)
```

## Are there differences in activity patterns between weekdays and weekends?
For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

1.  Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

2.  Make a panel plot containing a time series plot (i.e. type="1") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

### Results

1.  New factor variable with two levels - "weekday" and "weekend".
```{r Add new day of week column, echo=TRUE}
activityDataImputeDow <- activityDataImpute
activityDataImputeDow$day_of_week <- weekdays(activityDataImputeDow$datetime)
activityDataImputeDow$day_of_week_type <- as.factor(ifelse(activityDataImputeDow$day_of_week %in% c("Saturday", "Sunday"), "weekend", "weekday"))

tail(activityDataImputeDow, 10)

```
2.  Time series plot (i.e. type="1") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)

```{r Average Number of Steps for each 5 minute interval by weekday and weekend, echo=TRUE}
meanStepsPerIntervalDayOfWeekType <- activityDataImputeDow[,c("steps", "interval", "day_of_week_type")] %>%
                              filter(!is.na(steps)) %>%
                              group_by(interval, day_of_week_type) %>%
                              summarize(mean_steps = mean(steps)) %>%
                              print

```
Use ggplot for the making the time series plot.
```{r Time Series Plot Average Number of Steps for each 5 minute interval by day of week type, echo=TRUE}
g <- ggplot(activityDataImputeDow, aes(x =interval, y=steps, colour = day_of_week_type))
g <- g + geom_line() +
      facet_grid(day_of_week_type ~ .) +
      guides(fill = F) +
      xlab("Interval") +
      ylab("Average Number of Steps") +
      ggtitle('Time Series Plot of the Average Number of Steps During the Weekday and Weekend') +
      theme(plot.title = element_text(hjust = 0.5), legend.position = "none")
print(g)
```

The plot shows that filling in the missing data with the average steps per interval increased the number of data points and there are more steps being taken during around the 600 and 900 intervals (6:00 AM to 9:00 AM ) during the weekdays.  This is probably due to the reason that people are waking up earlier during the weekday and going to work as compared to sleeping in durinng the weekend.
